---
title: "M. capitata genome v1 Ordination Analysis"
author: "Timothy Stephens"
date: "02/03/2023"
output: 
  html_document:
    keep_md: yes
---



# Setup

Setup R env. Load packages and set default image export formats, size and resolution.

```{r setup}
knitr::opts_chunk$set(echo = TRUE,
                      fig.height = 8, 
                      fig.width = 12, 
                      dev = c("png", "pdf"),
                      dpi = 1000)
library(lubridate)
#library(tidyverse)
library(seacarb)
library(matrixStats)
library(vegan)
library(lme4)
library(ape)
library(emmeans)
library(gridExtra)
library(multcompView)
library(plotrix)
library(reshape2)
library(ggpubr)
library(sva, warn.conflicts = FALSE)
library(mixOmics)
library(ggforce)
library(plyr)
library(dplyr)
library(plotly)
library(gapminder)
library(htmlwidgets)
options(scipen = 999) #Prevent scientific notation
set.seed(54321)

#source("ggbiplot/R/ggbiplot.r")
```



Plot settings.
```{r plot_settings}
options(repr.plot.width = 2, repr.plot.height = 3)
label.decimal.rounding <- 1
plot.text.size  <- 12
plot.title.size <- 10

Colors <- c("T1-Amb" = "#00ebf7", "T3-Amb" = "#01ff00", "T5-Amb" = "#006600", "Field" = "#660099", "T1-HiT" = "#fad700", "T3-HiT" = "#ff7c00", "T5-HiT" = "#ff0000")
```



# Functions

Functions for plotting and analyzing data.

```{r func-run_PERMANOVA}
#' Run PERMANOVA analysis
#' 
#' @param data expression/accumulation dataframe (rownames=gene_names; colnames=sample_names)
#' @param samplesInfo sample_names metadata (required: "TimePoint", "Treatment", and "Tank")
#' @param out.prefix prefix of output csv file with PERMANOVA results
#' @param permutations number of permutations to perform (default: 999)
#' @return PERMANOVA results from vegan::adonis2
run_PERMANOVA <- function(data, samplesInfo, out.prefix, permutations=999){
  # Transpose df (samples as rownames)
  t.data <- t(data)
  t.data <- as.data.frame(t.data)

  # Identify factors
  #  - Will only return overlap between data and samplesInfo
  #  - Make the "by" column (which was used for merging) rownames.
  t.data.sampleInfo <- merge(samplesInfo, t.data, by = 0) %>%
    tibble::column_to_rownames(var = "Row.names")
  t.data.sampleInfo <- t.data.sampleInfo[row.names(t.data), ]
  
  # Use square root or proportions to minimize influence of most abundant groups
  t.data.sqrt <- sqrt(t.data)
  
  # Create a dissimilarity matrix (vegan)
  t.data.dist <- vegdist(t.data.sqrt, method='bray')
  
  # Run perMANOVA (vegan)
  mod.data <- adonis2(t.data.dist ~  TimePoint * Treatment, 
                      data = t.data.sampleInfo, 
                      permutations = permutations, 
                      strata = t.data.sampleInfo$Tank, 
                      method = 'bray')
  
  # Write PERMANOVA results to file
  write.csv(mod.data, paste(out.prefix,".PERMANOVA_results.csv", sep=''))
  
  # Return PERMANOVA results from function
  return(mod.data)
}
```



```{r func-plot_PCA}
#' Plot PCA
#' 
#' @param data expression/accumulation dataframe (rownames=gene_names; colnames=sample_names)
#' @param samplesInfo sample_names metadata (required: "Condition", "Treatment", and "Plug.ID")
#' @param out.prefix prefix of output html file with PCA plot
#' @param plot.title plot title
#' @return PCA plot as a ggplot2 object
plot_PCA <- function(data, samplesInfo, out.prefix, plot.title){
  # Transpose df (samples as rownames)
  t.data <- t(data)
  t.data <- as.data.frame(t.data)
  
  # Run PCA
  pca.out <- prcomp(t.data,
                    center = FALSE,
                    scale. = FALSE)
  pca.out.summary <- summary(pca.out)
  #print(pca.out.summary)
  
  # Create plotting dataframe by pulling out axes and joining to metadata
  PC1.df <- data.frame(pca.out$x[,1])
  names(PC1.df)[1] <- 'PC1'
  
  PC2.df <- data.frame(pca.out$x[,2])
  names(PC2.df)[1] <- 'PC2'
  
  # Merge PCA results with metadata sampleInfo file (used for easy access to variables for plotting)
  MyMerge <- function(x, y){
    return(merge(x, y, by=0, all.x=F, all.y= F) %>% 
             tibble::column_to_rownames(var = "Row.names"))
  }
  pca.data <- Reduce(MyMerge, list(samplesInfo, PC1.df, PC2.df))
  #print(pca.data)
  
  # Calculate the percent of variance explained by first two axes and format for plot labels
  PCA.axis1.var <- pca.out.summary$importance[2,1]
  PCA.axis2.var <- pca.out.summary$importance[2,2]
  PCA.axis1.var.label <- paste("PC1 (",round(PCA.axis1.var*100,label.decimal.rounding),"%)", sep='')
  PCA.axis2.var.label <- paste("PC2 (",round(PCA.axis2.var*100,label.decimal.rounding),"%)", sep='')
  
  # Plot PCA
  pca.plot <- ggplot(pca.data, aes(x = PC1, y = PC2)) +
              geom_point(aes(color = Condition, shape = Treatment, size = 2)) +
              scale_color_manual(values = Colors) +
              geom_mark_ellipse(aes(color = Condition),
                                expand = unit(1,"mm")) +
              labs(title=plot.title,
                   x=PCA.axis1.var.label,
                   y=PCA.axis2.var.label) +
              geom_text(label=pca.data$Plug.ID,
                        check_overlap=F) +
              theme(text       = element_text(family="Helvetica", size = plot.text.size),
                    plot.title = element_text(family="Helvetica", size = plot.title.size, hjust = 0.5),
                    aspect.ratio=1)
  
  # Write PCA plot as interactive HTML file
  saveWidget(suppressWarnings(ggplotly(pca.plot)), file=paste(out.prefix,".PCA.html", sep=''))
  
  # Return PCA plot from function
  return(pca.plot)
}
```



```{r func-plot_PCoA}
#' Plot PCoA
#' 
#' @param data expression/accumulation dataframe (rownames=gene_names; colnames=sample_names)
#' @param samplesInfo sample_names metadata (required: "Condition", "Treatment", and "Plug.ID")
#' @param out.prefix prefix of output html file with PCoA plot
#' @param plot.title plot title
#' @return PCoA plot as a ggplot2 object
plot_PCoA <- function(data, samplesInfo, out.prefix, plot.title){
  # Transpose df (samples as rownames)
  t.data <- t(data)
  t.data <- as.data.frame(t.data)
  
  # Identify factors
  #  - Will only return overlap between data and samplesInfo
  #  - Make the "by" column (which was used for merging) rownames.
  t.data.sampleInfo <- merge(samplesInfo, t.data, by = 0) %>%
    tibble::column_to_rownames(var = "Row.names")
  
  # Use square root or proportions to minimize influence of most abundant groups
  t.data.sqrt <- sqrt(t.data)
  
  # Create a dissimilarity matrix (vegan)
  t.data.dist <- vegdist(t.data.sqrt, method='bray')
  
  # Run PCoA (vegan)
  pcoa <- wcmdscale(t.data.dist, k = 2, eig = TRUE, add = "cailliez")
  
  # Create plotting dataframe by pulling out axes and joining to metadata
  axis1 <- pcoa$points[,1]
  axis1.df <- data.frame(axis1)
  axis2 <- pcoa$points[,2]
  axis2.df <- data.frame(axis2)
  
  # Merge PCA results with metadata sampleInfo file (used for easy access to variables for plotting)
  MyMerge <- function(x, y){
    return(merge(x, y, by=0, all.x=F, all.y= F) %>% 
             tibble::column_to_rownames(var = "Row.names"))
  }
  pcoa.data <- Reduce(MyMerge, list(samplesInfo, axis1.df, axis2.df))
  #print(pcoa.data)
  
  # Calculate the percent of variance explained by first two axes.
  pcoa.axis1.var <- sum((as.vector(pcoa$eig)/sum(pcoa$eig))[1])
  pcoa.axis2.var <- sum((as.vector(pcoa$eig)/sum(pcoa$eig))[2])
  pcoa.axis1.var.label <- paste("PCoA 1 (",round(pcoa.axis1.var*100,label.decimal.rounding),"%)", sep='')
  pcoa.axis2.var.label <- paste("PCoA 2 (",round(pcoa.axis2.var*100,label.decimal.rounding),"%)", sep='')
  
  pcoa.plot <- ggplot(pcoa.data, aes(x = axis1, y = axis2)) +
    geom_point(aes(color = Condition, shape = Treatment, size = 2)) +
    scale_color_manual(values = Colors) +
    geom_mark_ellipse(aes(color = Condition),
                      expand = unit(1,"mm")) +
    labs(title=plot.title,
         x=pcoa.axis1.var.label,
         y=pcoa.axis2.var.label) +
    geom_text(label=pcoa.data$Plug.ID,
              check_overlap=F) +
    theme(text       = element_text(family="Helvetica", size = plot.text.size),
          plot.title = element_text(family="Helvetica", size = plot.title.size, hjust = 0.5),
          aspect.ratio=1)

  # Write PCoA plot as interactive HTML file
  saveWidget(suppressWarnings(ggplotly(pcoa.plot)), file=paste(out.prefix,".PCoA.html", sep=''))
  
  # Return PCoA plot from function
  return(pcoa.plot)
}
```



```{r func-plot_PLS-DA}
#' Plot PLS-DA
#' 
#' @param data expression/accumulation dataframe (rownames=gene_names; colnames=sample_names)
#' @param samplesInfo sample_names metadata (required: "Group", "Condition", "Treatment", and "Plug.ID"). Where "Group" is Genotype+TimePoint+Treatment
#' @param out.prefix prefix of output html file with PLS-DA plot
#' @param plot.title plot title
#' @return PLS-DA plot as a ggplot2 object
plot_PLSDA <- function(data, samplesInfo, out.prefix, plot.title){
  # Transpose df (samples as rownames)
  t.data <- t(data)
  t.data <- as.data.frame(t.data)
  
  # Set "group" as vector
  samplesInfo <- samplesInfo %>% filter(row.names(.) %in% colnames(data))
  sample.info <- samplesInfo$Group
  sample.info <- as.factor(sample.info)

  # Determine the number of components to retain (ncomp -> K−1, where K is the number of classes)
  # In this case the components are the treatemnt "Groups" i.e., 
  #      - MC-289_Field MC-289_T1-Amb MC-289_T1-HiT MC-289_T3-Amb MC-289_T3-HiT MC-289_T5-Amb MC-289_T5-HiT
  # Total 7 levels (6 ncomp's)
  sample.class.number <- length(levels(sample.info))
  k <- sample.class.number-1
  #print(paste("Running with k=",k, sep=''))
  
  # Check data for correct dimensions
  #print(summary(sample.info))
  #print(length(sample.info))
  #print(dim(t.data))
  
  # Create PLSDA object
  plsda <- splsda(t.data, sample.info, ncomp = k, scale = FALSE, near.zero.var = TRUE) 
  
  # Pull PLS-DA 1 & 2
  plsda.df <- data.frame(plsda$variates$X[,1:2])
  names(plsda.df)[1] <- 'PLSDA1'
  names(plsda.df)[2] <- 'PLSDA2'
  
  # Merge PCA results with metadata sampleInfo file (used for easy access to variables for plotting)
  MyMerge <- function(x, y){
    return(merge(x, y, by=0, all.x=F, all.y= F) %>% 
             tibble::column_to_rownames(var = "Row.names"))
  }
  plsda.data <- Reduce(MyMerge, list(samplesInfo, plsda.df))
  #print(plsda.data)
  
  # Calculate the percent of variance explained by first two axes.
  plsda$prop_expl_var$X
  plsda.axis1.var <- plsda$prop_expl_var$X[1]
  plsda.axis2.var <- plsda$prop_expl_var$X[2]
  plsda.axis1.var.label <- paste("PLS-DA 1 (",round(plsda.axis1.var*100,label.decimal.rounding),"%)", sep='')
  plsda.axis2.var.label <- paste("PLS-DA 2 (",round(plsda.axis2.var*100,label.decimal.rounding),"%)", sep='')
  
  # Plot initial PLSDA object
  plsda.plot <- ggplot(plsda.data, aes(x = PLSDA1, y = PLSDA2)) +
    geom_point(aes(color = Condition, shape = Treatment, size = 2)) +
    scale_color_manual(values = Colors) +
    geom_mark_ellipse(aes(color = Condition),
                      expand = unit(1,"mm")) +
    labs(title=plot.title,
         x=plsda.axis1.var.label,
         y=plsda.axis2.var.label) +
    geom_text(label=plsda.data$Plug.ID,
              check_overlap=F) +
    theme(text       = element_text(family="Helvetica", size = plot.text.size),
          plot.title = element_text(family="Helvetica", size = plot.title.size, hjust = 0.5),
          aspect.ratio=1)
  
  # Write PLS-DA plot as interactive HTML file
  saveWidget(suppressWarnings(ggplotly(plsda.plot)), file=paste(out.prefix,".PLS-DA.html", sep=''))
  
  # Return PLS-DA plot from function
  return(plsda.plot)
}
```





# Sample metadata

Load sample metadata to use for all xpression/accumulation datasets.
```{r load_data}
# Make "Sample.ID" row names
samplesInfo <- read.table("samples_info.txt", h = T, row.names = NULL, sep = "\t", check.names = F) %>% 
  tibble::column_to_rownames(var = "Sample.ID")

# Convert to character variables (so they are treated as descrete and not continuious) and factors.
samplesInfo$Treatment <- as.factor(as.character(samplesInfo$Treatment))
samplesInfo$TimePoint <- as.factor(as.character(samplesInfo$TimePoint))
samplesInfo$Tank      <- as.factor(as.character(samplesInfo$Tank))

samplesInfo
```





# Analyze proteomic data

```{r proteomic_data}
# Load expression data:
#   - Make "Names" column rownames
#   - Make "NA" values "0"
proteomic.data <- read.table("Mcapitata_V1_proteomic_data.tsv", h = T, row.names = NULL, sep = "\t", check.names = F) %>% 
  tibble::column_to_rownames(var = "Name") %>% 
  replace(is.na(.), 0)
dim(proteomic.data)

# Set file names
out.prefix <- "Mcapitata_V1_proteomic_data"

run_PERMANOVA(proteomic.data, samplesInfo, out.prefix, permutations=how(nperm=190, minperm=190))
proteomic.PCA       <- plot_PCA(     proteomic.data, samplesInfo, out.prefix, "Proteomic data (PCA)")
proteomic.PCoA      <- plot_PCoA(    proteomic.data, samplesInfo, out.prefix, "Proteomic data (PCoA)")
proteomic.PLSDA     <- plot_PLSDA(   proteomic.data, samplesInfo, out.prefix, "Proteomic data (sPLS-DA)")
```





# Analyze transcriptomic data

```{r Transcriptomic_data}
# Load expression data:
#   - Make "Names" column rownames
#   - Make "NA" values "0"
#   - Ignore rows with a combined (summed) TPM across samples <=100
transcriptomic.data <- read.table("Mcapitata_V1_transcriptomic_data.tsv.TPM.txt", h = T, row.names = NULL, sep = "\t", check.names = F) %>% 
  tibble::column_to_rownames(var = "Name") %>% 
  replace(is.na(.), 0) %>% 
  filter(rowSums(.) > 100)
dim(transcriptomic.data)

# Set file names
out.prefix <- "Mcapitata_V1_transcriptomic_data"

run_PERMANOVA(transcriptomic.data, samplesInfo, out.prefix)
transcriptomic.PCA       <- plot_PCA(     transcriptomic.data, samplesInfo, out.prefix, "Transcriptomic data (PCA)")
transcriptomic.PCoA      <- plot_PCoA(    transcriptomic.data, samplesInfo, out.prefix, "Transcriptomic data (PCoA)")
transcriptomic.PLSDA     <- plot_PLSDA(   transcriptomic.data, samplesInfo, out.prefix, "Transcriptomic data (sPLS-DA)")
```





# Analyze transcriptomic data that overlaps with proteomic data

```{r Transcriptomic-Proteomic-Overlap_data}
# Load expression data:
#   - Make "Names" column rownames
#   - Make "NA" values "0"
#   - Ignore rows with a combined (summed) TPM across samples <=100
#   - **Keep only rows in proteomic data**
trans_prot_overlap.data <- transcriptomic.data %>%
  filter(row.names(.) %in% rownames(proteomic.data))
dim(trans_prot_overlap.data)

# Set file names
out.prefix <- "Mcapitata_V1_transcriptomic-proteomic_overlap_data"

run_PERMANOVA(trans_prot_overlap.data, samplesInfo, out.prefix)
trans_prot_overlap.PCA       <- plot_PCA(     trans_prot_overlap.data, samplesInfo, out.prefix, "Transcripts with proteomic evidence (PCA)")
trans_prot_overlap.PCoA      <- plot_PCoA(    trans_prot_overlap.data, samplesInfo, out.prefix, "Transcripts with proteomic evidence (PCoA)")
trans_prot_overlap.PLSDA     <- plot_PLSDA(   trans_prot_overlap.data, samplesInfo, out.prefix, "Transcripts with proteomic evidence (sPLS-DA)")
```





# Combined plots

```{r combined_plot_main}
ggarrange(proteomic.PCoA, transcriptomic.PCoA, trans_prot_overlap.PCoA,
          ncol = 2, nrow = 2,
          labels = "AUTO",
          font.label = list(size = 24, color = "black", face = "bold", family = "Helvetica"),
          hjust = -3,
          common.legend = TRUE,
          legend = "right")
```


```{r combined_plot_supp}
ggarrange(proteomic.PLSDA, transcriptomic.PLSDA, trans_prot_overlap.PLSDA,
          proteomic.PCA,   transcriptomic.PCA,   trans_prot_overlap.PCA,
          ncol = 3, nrow = 2,
          labels = "AUTO",
          font.label = list(size = 24, color = "black", face = "bold", family = "Helvetica"),
          hjust = -0.2,
          common.legend = TRUE,
          legend = "right")
```





# Session Info

```{r ressionInfo}
sessionInfo()
```
